### 2013-04-22 article.tex
Further proof reading of manuscript
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. In contrast there remains a
steep learning curve in the subsequent computational steps required to process
raw sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. Genomer simplifies updating a
genome scaffold by allowing a human-readable YAML format file to be edited
instead of large sequence files. Genomer is written as a command line tool and
is an effort to make the manual, and increasingly common, process of genome
scaffolding more robust and reproducible. Extensive documentation and video
tutorials are available at http://next.gs.

@@

The decreasing costs and increasing diversity of high-throughput sequencing
methods is making genome analysis a common method to tackle unresolved
questions in microbiology \cite{loman2012b}. Sequencing can produce
$10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
larger contigs then scaffolded into larger megabase fragments or complete
chromosomes. These larger DNA sequences are then annotated with genomic
features such as protein-coding genes. Each of these required steps can however
produce imperfect results and remains the subject of active research
\cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
sequence and corresponding annotations may still improve upon the results of
automated methods.

Manually editing a genome sequence or annotation is a non-trivial problem
requiring effort on a researcher's part. The large size of FASTA sequence files
or general feature format (GFF) annotation files makes changing these in a text
editor difficult. Tasks such as organising project files, adding additional
sequences to an assembly, updating incorrect annotations or preparing files for
database deposition all divert effort from the main goal of analysing a genome.
As a result there is a steep learning curve for researchers embarking on genome
sequencing for the first time, while still requiring effort for experienced
researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer,'' to ease and simplify many of the
common steps in our genome project. While several graphical user interface
tools are available to work with genome projects \cite{tanaka2006,
wilkinson2002, lopez2011, carver2012, gordon1998}, Genomer instead provides a
simple command line interface. This allows scriptable automation and
reproducible results. We have open-sourced this tool under the MIT license and
here we describe its application.

@@

Genomer is written in the Ruby programming language \cite{ruby-lang,goto2010}
and tested against the 1.8.7 and 1.9.3 versions. Genomer is built upon the
Rubygems and Bundler Ruby package management libraries. Rubygems is a software
management system for Ruby which allows automatic downloading and installation
of third-party software libraries from the rubygems.org website. Bundler is
software for determining which specific versions of the available third
libraries should be used.

@@

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Plugins
  can also be locked to specific versions to prevent backwards incompatible
  software changes breaking an existing workflow.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool built around plain text files. This allows for integration with common
  Linux tools tools such as GNU Make or git. This thereby allow reproducible
  scripting of the genome finishing process.}

@@

Genomer is written for the command-line to use in shells such as bash or zsh.
Command line tools are simple to automate using scripting, allowing
reproducible genome finishing and sharing genome projects between researchers.
For instance, during our microbial genome project we automated Genomer using
GNU Make build files, allowing the finishing steps to be repeated automatically
using the \verb+make+ command. This is an advantage of command-line tools
versus GUI-based tools, is that the former are simpler to automate and share
steps with other researchers. Therefore Genomer may be of particular interest
to bioinformaticians who prefer automated approaches and scripting on the
command line.

Additionally as Genomer uses plain text it is easy to store versions of the
project using a revision control system such as git \cite{git-scm}. This allows
tracking changes to the project and reverting errors back to earlier versions.
Example versioned build files from our genome projects can be found on GitHub
for a simple plasmid \cite{plasmid-github} and a more complex genome
\cite{genome-github} sequence.

We have released Genomer as open source on GitHub \cite{genomer-github} and
have prepared documentation and video tutorials at http://next.gs. As we have
used this tool extensively to simplify our own genome projects we believe
Genomer may also be useful to others in the field. We will illustrate the
possible application of Genomer with an example use case.

@@

  \item{A Genomer project is organised around a set of already assembled and
  annotated contigs. For example we assemble our reads into contigs using the
  A5 pipeline \cite{tritt2012} and and then subsequently annotate them using
  the Integrated Microbial Genomes resource \cite{markowitz2006}. We then use
  these contigs as a starting point to build a draft genome sequence in
  Genomer.}

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs in the
  scaffold. Any unresolved gap regions can also be specified in this file.
  Paired-end sequencing or comparison to reference genomes are a source of
  contig orientation and order. In our genome projects we determine contig
  order by aligning contigs to a reference genome using nucmer
  \cite{kurtz2004}. A detailed overview of the scaffolder file format can be
  found on the Genomer website.}

  \item{The \verb+genomer view+ command can be used to generate files of the
  assembled scaffold. These may include the FASTA file of the assembly or a
  GFF3 file of updated annotations corresponding to the assembled genome
  sequence. These files can then be used for downstream analyses of the
  sequence.}

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and \emph{in silico} analysis. Changes to the
  scaffold file automatically propagate updates in the generated files when the
  \verb+genomer view+ command is rerun.}

  \item{When researchers are satisfied with the status of their draft the
  required files for submission to GenBank can be generated: assembled FASTA,
  contig FASTA, annotation tables, and scaffold layout in AGP. Additional
  options are used to add the required prefix to gene annotation IDs and begin
  locus ID numbering at the origin of the sequence. Genomer does not produce
  the .asn file which is submitted to GenBank, but instead the files that may
  be used as the input to tbl2asn or sequin. }

@@

Genomer was written to satisfy our needs for building and finishing a draft
microbial genomes less than 10MBp in size. Draft sequence assembly takes place
in memory and therefore building very large sequences in memory may limit
performance. We point out however that sizes of even 3GBp should fit in the
memory of a modern laptop or desktop computer.

Genomer simplifies the process of generating the annotation table file required
for submission to GenBank where only gene entries are required. Specifically a
user need only provide a GFF3 file of annotations and Genomer will generate the
additional encoded protein features. This however assumes no intron/exon
structure and therefore users wishing to submit annotations with alternative
splicing cannot take advantage of this Genomer feature.

>>>>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. Yet there remains a steep
learning curve in the subsequent computational steps required to process raw
sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. This software also simplifies
updating a genome scaffold by allowing a human-readable YAML format file to be
edited instead of large sequence files. Genomer is written as a command line
tool and is an effort to make the manual, and increasingly common, process of
genome scaffolding more robust and reproducible. Extensive documentation and
video tutorials are available at http://next.gs.

@@

The decreasing costs and increasing diversity of high-throughput sequencing
methods is making genome analysis a common method to tackle unresolved
questions in microbiology \cite{loman2012b}. Sequencing can produce
$10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
larger contigs which are then scaffolded into larger megabase fragments or
complete chromosomes. These larger DNA sequences are then annotated with
genomic features such as protein-coding genes. Each of these required steps can
however produce imperfect results and remains the subject of active research
\cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
sequence and corresponding annotations may still improve upon the results of
automated methods.

Manually editing a genome sequence or annotation is a non-trivial problem
requiring effort on a researcher's part. The large size of FASTA sequence files
or general feature format (GFF) annotation files makes changing these in a text
editor difficult. Tasks such as organising project files, adding additional
sequences to an assembly, updating incorrect annotations or preparing files for
database deposition all divert effort from the main goal of analysing a genome.
As a result there is a steep learning curve for researchers embarking on genome
sequencing for the first time, while still requiring effort from experienced
researchers.

@@

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer,'' to ease and simplify many of the
common steps in our genome project. While several graphical user interface
tools are available to work with genome projects, including G-InforBio,
Genquire, Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon1998}. Genomer provides a simple command line
interface which allows scriptable automation and reproducible results. We made
this software open-source under the MIT license and here we describe its
application.

@@

Genomer is written in the Ruby programming language (1.8.7 and 1.9.3)
\cite{ruby-lang,goto2010} and built upon the Rubygems and Bundler Ruby package
management libraries. Rubygems is a software management system for Ruby which
allows automatic downloading and installation of third-party software libraries
from the rubygems.org website. Bundler is software for determining which
specific versions of the available third libraries should be used.

@@

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Install
  plugins may also be locked to specific versions to prevent backwards
  incompatible software changes breaking an existing workflow.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool around plain text files. This allows for integration with common Linux
  tools tools such as GNU Make or git thereby allowing reproducible scripting
  of the genome finishing process.}

@@

Genomer is written as a command-line tool for use in shells such as bash or
zsh. The advantage of command-line tools over GUI-based tools, is that the former
are simpler to automate and allow the sharing of steps with other researchers
working on related projects. For instance, during our microbial genome project
we automated Genomer using GNU Make build files, allowing the finishing steps
to be repeated automatically using the \verb+make+ command. Genomer may
therefore be of particular interest to bioinformaticians coordinating in groups
on the same genome project and those who prefer automated approaches and
scripting on the command line.

As Genomer uses plain text files to manage the draft genome sequence it is easy
to store versions of the project using a revision control system such as git
\cite{git-scm}. This allows tracking changes to the project and reverting
errors back to earlier versions. Example versioned build files from our genome
projects can be found on GitHub for a simple plasmid \cite{plasmid-github} and
a more complex genome \cite{genome-github} sequence.

We have released Genomer as open source software on GitHub
\cite{genomer-github} and have prepared documentation and video tutorials at
http://next.gs. We have used this tool extensively to simplify our own genome
projects and believe Genomer may also be useful to others in the field. We will
illustrate the possible application of Genomer with the following example use
case.

@@

  \item{A Genomer project is organised around a set of already assembled and
  annotated contigs. In our microbial genome projects we assemble our reads
  into contigs using the A5 pipeline \cite{tritt2012} and then subsequently
  annotate them using the Integrated Microbial Genomes resource
  \cite{markowitz2006}. We then use these contigs as a starting point to build
  a draft genome sequence in Genomer.}

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs and
  unresolved gap regions in the scaffold. Paired-read sequencing or comparison
  to reference genomes can be used as a source of contig orientation and order.
  As an example, in our own genome projects we determine contig order by
  aligning contigs to a reference genome using nucmer \cite{kurtz2004}. A
  detailed overview of the scaffolder file format can be found on the Genomer
  website.}

  \item{The \verb+genomer view+ command is used to generate files of the
  assembled scaffold, such as the FASTA file of the assembly or a GFF3 file of
  annotations locations on the assembled genome sequence. These files can then
  be used for subsequent analyses of the sequence.}

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and \emph{in silico} analysis. Genomer allows
  changes in the scaffold file to automatically propagate to the required
  changes in the generated files when the \verb+genomer view+ command is
  rerun.}

  \item{The required files for submission to GenBank may be generated when
  researchers are satisfied with the status of their draft. These files include
  the assembled FASTA file, contig FASTA file, annotation table file, and AGP
  scaffold file. Additional options can be used to add the required prefix to
  gene annotation IDs and begin locus ID numbering at the origin of the
  sequence. Genomer does not produce the .asn file which is submitted to
  GenBank, but instead the files that may be used as the input to tbl2asn or
  sequin provided by GenBank.}

@@

Genomer was written to satisfy our needs for building and finishing megabase
sized draft genomes where assembly of the draft sequence takes place in memory.
This makes Genomer well-suited for microbial genome projects where
approximately 99\% of microbial genomes sequenced to date are less than 10MBp.
The Assembly of very large sequences may however limit performance when the
size of the sequence exceeds available. We however anticipate that genome sizes
of even 3GBp will fit in the memory of a modern laptop or desktop computer.

One features of Genomer is the simplification of generating the annotation
table file required for submission to GenBank. Genomer requires only a GFF3
file containing gene-type entries be required. Genomer is able to generate the
additional encoded protein features from these entries. One caveat to this
however is that the assumption there is no intron/exon structure and therefore
users wishing to submit annotations with alternative splicing cannot take
advantage of this Genomer feature.
